Species motif participation provides unique information about species risk of extinction

### 2022-06-16 manuscript/supplement_motifs.tex
added globalweb ref
Updates from Overleaf

\documentclass[12pt]{article} 
\usepackage{amsmath} 
\usepackage[dvips]{graphicx}
\usepackage{multirow} 
\usepackage{geometry} 
\usepackage{pdflscape}
\usepackage[labelfont=bf]{caption} 
\usepackage{setspace}
\usepackage[running]{lineno} 
% \usepackage[numbers,sort]{natbib}
\usepackage[round]{natbib} 
\usepackage{array}
\usepackage[table]{xcolor}
\usepackage{xr}

@@

\title{Motif roles provides unique information about species risk of extinction\\ \medskip Supplementary Information}

\author{Anna \r{A}kesson$^{1\dagger}$, Alyssa R. Cirtwill$^{2}$, Kate L. Wootton$^{3}$, Anna Ekl\"{o}f$^{1}$} 
\date{\small$^1$Department of Theoretical Biology, Chemistry, and Physics\\ 
Link\"{o}ping University\\
Link\"{o}ping, Sweden\\
\medskip
\small$^2$Department of Agricultural Sciences\\
University of Helsinki\\
Helsinki, Finland\\
\medskip
\small$^3$ BioFrontiers Institute\\
University of Colorado, Boulder\\
U.S.A.
\medskip
$^\dagger$ Corresponding author:\\
}

@@

\section*{Table of contents}
    \begin{spacing}{1.0}

@@

    \end{enumerate}    
\end{spacing}
\clearpage

@@

        We use Bayesian network modelling to calculate the persistence of each species in our simulated networks. 
        A Bayesian network represents a graphical model for probabilistic relationships among a set of variables. 
        In our ecological setting, the nodes of the network represent species and links represent feeding interactions between species.
        In network-theory terms, the species are Bernoulli random variables, and the links represent conditional dependencies among these variables. 
        Each node has a set of local probability dependencies, conditional on the state of the parent nodes. 
        Since the nodes are species, their state can be either extant or extinct.
        The extinction probability of each species is a function of the state of its resources and also depends on a baseline probability of extinction ($\pi$) representing risk of extinction due to factors unrelated to network structure.
        In our baseline scenario, basal resources had a baseline extinction probability of $\pi = 0.1$. 
        We also simulated a set of scenarios where basal resources had a higher baseline probability of extinction ($\pi$). 
        These scenarios reflect threats to basal resources such as increased temperatures resulting in droughts, making some species more vulnerable to extinction.
        $\pi$ ranged between $0.1-0.5$, in steps of $0.08$. 
        The highest disturbance level, $\pi = 0.5$, corresponds to basal species having a 50\% risk of going extinct. 
        Consumer species retained $\pi=0.1$ in all cases.

@@

        As consumer species depend on their resources, we start by determining the status of each primary producer in the network.
        Systematically, we use Equation \ref{betafunc} to calculate each species' probability of extinction $P(\lnot i|f)$, and then perform Bernoulli trials to determine whether the species goes extinct or not. 
        Imagine a basal species \textit{A} with extinction probability $P(\lnot A) = \pi$ where $\pi = 0.1$. We draw a random number $r$ from a uniform distribution between $0-1$. The basal species would be considered extinct if $r < 0.1$, and extant otherwise.

@@

        We simulated a suite of food webs based on the probabilistic niche model, which assigns predator-prey links based on the body-mass ratios between individuals of different species~\citep{Williams2000,Delmas2017}. The meso-scale structure of niche-model networks closely mimics that of empirical food webs~\citep{Stouffer2007}. To ensure that we captured a variety of realistic community sizes and structures, we generated networks ranging between 50 and 100 species (in steps of 10) with connectance values between 0.02 and 0.2 (in steps of 0.02). The range of network sizes was chosen to reflect moderately well-sampled empirical webs while working within our computational limits, while the range of connectance values was chosen to cover that observed in most empirical food webs~\citep{Dunne2002e}. We generated a total of 100 networks with each combination of parameters, for a total of 6000 networks. All networks were generated using the function "nichemodel" within the Julia language package \emph{BioEnergeticFoodWebs}~\citep{bioenergeticfw,Delmas2017}. If a simulated network contained any disconnected species (species without predators or prey) or disconnected components (a group of species connected among themselves but not to the rest of the network), the network was rejected and a new network simulated. Finally, networks where the path lengths between each species and a basal resource could not be resolved (i.e., trophic levels were undefined) were rejected and new networks simulated.

        After generating the network structure, we simulated community dynamics using the function ``simulate'' from the Julia language package \emph{BioEnergeticFoodWebs}~\citep{bioenergeticfw,Delmas2017}. This function uses Lotka-Volterra predator-prey models including density dependence and type 2 functional responses for all species (please see~\citet{Delmas2017} for full details).
        All non-basal species were designated as vertebrates to ensure a good match between metabolic and predator-prey body-mass ratio values. Metabolic rates in the Lotka-Volterra model are based on each species' body mass (i.e. mass of a single individual). We assigned relative body masses based on each species' trophic level, which was, in turn, calculated based on the food-web structure provided by the niche model. After basal species were assigned a body mass of 1, we used a predator-prey body-mass ratio of 3.065 to calculate the relative body masses of higher trophic levels. We selected this ratio based on the estimate for vertebrates (averaged across ecosystem and metabolic types) in~\citet{Brose2006}. We excluded reported body-mass ratios for invertebrates as these could include parasites and parasitoids, which are generally smaller than their prey, and because interactions among vertebrates are better represented in the food-web literature than interactions involving invertebrates.

        The persistence of each species in our simulated networks also depends on its population biomass. 
        We randomly assigned initial population biomasses (i.e. cumulative biomass across all individuals of a species) for each species from a uniform distribution [0,1]. Note that population biomasses and individual body masses are not calculated on the same scale. We then simulated community dynamics for 1000 time steps to obtain an `equilibrium' community. To ensure that species did not `recover' from unrealistically low biomasses during the simulation, we considered a species extinct if it dropped below an arbitrary threshold biomass of 1$\times10^{-5}$. When simulating initial (i.e. pre-disturbance or equilibrium) dynamics, we rejected any network where one or more species dropped below this biomass threshold. Consumers were assumed to have no preferences such that the consumption rate $w_{ij}$ of predator $i$ eating prey $j$ is equal to $\frac{1}{n}$, where $n$ is the number of prey for predator $i$. If the network did not retain all species for 1000 time steps, a new set of initial population biomasses was applied and the simulation repeated.
        If a set of biomasses where all species persist still had not been reached after 100 sets of randomly-assigned initial biomasses, we discarded the network and simulated another to replace it.

@@

        A Bayesian food web depicts probabilistic relationships among a set of species where each species' probability of persistence depends upon the probabilities of its resource species persisting~\citep{Jensen_Nielsen,Eklof2013}. 
        All persistence calculations therefore begin by determining the status of the basal resources (who do not depend on other species).
        Calculations continue in a strictly bottom-up manner with primary consumers (who depend only on basal resources), and so on up the network.

@@

        \begin{figure}
            \includegraphics[width=\textwidth]{figures/motif_profiles_participation_vs_empirical.eps}
            \caption{\textbf{A)} Proportions of acyclic motifs in the motif profiles of empirical and simulated networks. Mean proportions ($\pm$SD) in the simulated networks are indicated in black, with the range of proportions in the simulated networks indicated by black boxes. \textbf{B)} Proportions of acyclic motifs in the motif participation vectors of consumers in simulated and empirical networks.}
            \label{empirical_compare}
        \end{figure}

@@

    \begin{figure}[hb!]
      \includegraphics[width=\textwidth]{figures/heatmap_CS_BPcompare.eps}
       \caption{Heat maps of average consumer persistence in networks with varying connectance (y-axis) and network size (x-axis) at two levels of disturbance to basal resources. Persistence decreases with darker color in the heat map. When basal species are not disturbed, all species experience a baseline extinction risk of $\pi = 0.1$. When basal species are highly disturbed, their baseline extinction risk rises to $\pi = 0.5$ while the baseline risk of consumers is not changed.}
       \label{fig:heatmap_CS}
    \end{figure}

@@

% \section{SH: Persistence and motif roles}

%     \subsection*{Methods}

%         As one more level of detail, we can consider the frequency with which a species appears in each \emph{position} in each motif. 
%         Different positions are associated with different numbers of predators and prey, making the similarities and differences between motifs and degree more explicit.
%         The full set of these positions, across all motifs, is the species' \emph{motif role}~\citep{Stouffer2010,Cirtwill2017}.
%         As with motif participation, we normalized motif roles by dividing the number of times a species appeared in each position by the total number of times it appeared in all positions~\citep{Baker2015,Cirtwill2017}.

%         [[Could add a PERMANOVA to show that species' roles overal affect their persistence. Haven't done so yet.]]

%         To test whether participating in different motif positions was related to a species' probability of persistence, we fit similar models to those for motif participation.
%         For each position, we fit a linear mixed-effect model relating persistence to the frequency of the position in a species' motif role, disturbance, and their interaction.
%         To account for differences in persistence across networks of different size and connectance, we included a random effect of the interaction between size and connectance.
%         All regressions were fit using the R~\citep{R} function `lmer' from the package \emph{lmerTest}~\citep{lmerTest}.

%     \subsection*{Results}

%     \begin{table}[h!]
%         \caption{Coefficients in a series of regressions relating probability of persistence to the frequency of a position in a species' motif role, disturbance, and their interaction. For ease of comparison between motifs, we did not center or scale predictors. Note, however, that some motifs and positions have higher frequencies than others.}
%         \label{tab:persistence_vs_positions}
%         \centering
%         \begin{tabular}{l | c c c c}
%         Position    & Intercept & Disturbance   & Position  & Interaction \\
%         \hline
%         Chain: top  & 0.880 &   -0.985 & -0.00544 & -0.332 \\
%         Chain: bottom   & 0.883 &   -1.07 & -0.0510 &   0.647 \\
%         Chain: middle   & 0.862 & -1.00 &   0.246 & -0.244 \\
%         \hline
%         Dir. Comp.: top & 0.872 &   -0.995  & 0.0549 &  -0.178 \\
%         Dir. Comp.: bottom & 0.886 &    -1.07 & -0.113 &    0.939 \\
%         \hline
%         Omnivory: top   & 0.872 &   -0.984 &    0.180 & -0.818 \\
%         Omnivory: middle & 0.873 & -1.00 &  0.119 & -0.324 \\
%         Omnivory: bottom & 0.882 & -1.06 &  -0.0449 &   0.845 \\
%         \hline
%         App. Comp.: top & 0.865 &   -0.966 &    0.140 & -0.520 \\
%         App. Comp.: bottom & 0.910 &    -1.12 & -0.100 &    0.323 \\
%         \hline
%         \end{tabular}
%     \end{table}

%     \begin{figure}
%         \centering
%         \includegraphics[width=0.75\textwidth]{figures/persistence_positions_Apparent.eps}
%         \caption{A species' probability of persistence varied with the frequency of different motif positions in its role. Here we show the relationships between persistence and the proportions of the three positions in the apparent competition motif at three levels of probability of extinction for basal resources (p=0.1, 0.26, and 0.42), indicated by line colour. Line style indicates the motif  position. }
%         \label{fig:apparent_positions}
%     \end{figure}

%     \begin{figure}
%         \centering
%         \includegraphics[width=0.75\textwidth]{figures/persistence_positions_Direct.eps}
%         \caption{A species' probability of persistence varied with the frequency of different motif positions in its role. Here we show the relationships between persistence and the proportions of the three positions in the direct competition motif at three levels of probability of extinction for basal resources (p=0.1, 0.26, and 0.42), indicated by line colour. Line style indicates the motif  position. }
%         \label{fig:direct_positions}
%     \end{figure}

%     \begin{figure}
%         \centering
%         \includegraphics[width=0.75\textwidth]{figures/persistence_positions_Three-species.eps}
%         \caption{A species' probability of persistence varied with the frequency of different motif positions in its role. Here we show the relationships between persistence and the proportions of the three positions in the three-species chain motif at three levels of probability of extinction for basal resources (p=0.1, 0.26, and 0.42), indicated by line colour. Line style indicates the motif  position. }
%         \label{fig:chain_positions}
%     \end{figure}

% \clearpage

% \section{SI: Motif roles and network properties}

%     \subsection*{Methods}

%         As with network motif profiles and motif participation, species' motif roles may vary with network size and connectance and with the species' in degree and trophic level.
%         To identify these relationships, we fit two linear regressions for each motif position.
%         The first related the frequency of the position in species' roles to network size, connectance, and their interaction.
%         The second related the frequency of the position to in-degree, STL, and their interaction.
%         All regressions were fit using the R~\citep{R} base function `lm'.

%     \subsection*{Results}
%         \subsection*{Degree and trophic level}

%             Positions in all three motifs were strongly related to in-degree and/or trophic level (Table~\ref{DT_tab}).
%             The top position in the omnivory motif increased with increasing degree across all trophic levels.
%             The frequency of the middle position increased with the interaction of degree and trophic level.
%             The bottom position was generaly rare, especially for species with high degree or high trophic level (this suggests that the bottom position is frequently a basal resource).

%             The frequency of the bottom position in the apparent competition motif decreased strongly with increasing degree, especially for species at a high trophic level.
%             The frequency of the top position increased strongly with increasing degree and was not strongly related to trophic level.

%             The frequency of the bottom position in the direct competition motif increased with increasing degree, especially for species at a high trophic level.
%             The bottom position was very rare, especially for species with high degree (also suggesting that this position is usually a basal resource).

%             The bottom position in the three-species chain was generally rare, especially for species with high degree or trophic level (again, this is likely to commonly be basal resources).
%             The middle position in the three-species chain did not vary strongly with degree and declined slightly with increasing trophic level.
%             The top position in the three-species chain increased with trophic level, degree, and their interaction.

%             \begin{table}[h!]
%                 \caption{Coefficients ($\beta$) and $p$-values for linear regressions of the frequency of each motif position against in-degree, STL, and their interaction.}
%                 \label{DT_tab}
%                 \footnotesize
%                 \begin{tabular}{c c | c c | cc | cc |}
%                 Motif & Position & \multicolumn{2}{c}{Degree} & \multicolumn{2}{c}{STL} & \multicolumn{2}{c}{Interaction} \\
%                 & & $\beta$ & $p$ & $\beta$ & $p$ & $\beta$ & $p$ \\
%                 \hline
%                 Chain & bottom & -1.47 $\times10^{-5}$ & 0.769 & -0.0218 & \textless0.001 & -0.00160 & \textless0.001 \\
%                 Chain & middle & 3.13$\times10^{-4}$ & \textless0.001 & -0.0353 & \textless0.001 & 1.85$\times10^{-5}$ & 0.275 \\
%                 Chain & top & -0.0172 & \textless0.001 & 0.0259 & \textless0.001 & 0.00919 & \textless0.001 \\
%                 \hline
%                 Omnivory & bottom & 6.14$\times10^{-4}$ & \textless0.001 & -0.00270 & \textless0.001 & -0.00112 & \textless0.001 \\
%                 Omnivory & middle & 3.13$\times10^{-4}$ & \textless0.001 & -0.0353 & \textless0.001 & 1.85$\times10^{-5}$ & \textless0.001 \\
%                 Omnivory & top & 0.00413 & \textless0.001 & 0.00381 & \textless0.001 & 9.96$\times10^{-5}$ & \textless0.001 \\
%                 \hline
%                 App. comp. & bottom & -4.74$\times10^{-4}$ & \textless0.001 & -0.663 & \textless0.001 & 0.00224 & \textless0.001 \\
%                 App. comp. & top & 2.12$\times10^{-5}$ & 0.183 & -0.112 & \textless0.001 & -4.47$\times10^{-4}$ & \textless0.001 \\
%                 \hline
%                 Dir. comp. & bottom & 8.12$\times10^{-4}$ & \textless0.001 & -5.57$\times10^{-3}$ & \textless0.001 & -1.37$\times10^{-3}$ & \textless0.001 \\
%                 Dir. comp. & top & -0.00405 & \textless0.001 & -0.0435 & \textless0.001 & 0.00260 & \textless0.001 \\
%                 \hline
%                 \end{tabular}
%                 \end{table}

%             \begin{figure}[h!]
%                 \centering
%                 \includegraphics[width=.75\textwidth]{figures/positions_byTL_Omnivory.eps}
%                 \caption{The frequencies of positions within the omnivory motif varied with degree, trophic level, and their interaction.}
%                 \label{omnivory_DT}
%             \end{figure}

%             \begin{figure}[h!]
%                 \centering
%                 \includegraphics[width=.75\textwidth]{figures/positions_byTL_Apparentcompetition.eps}
%                 \caption{The frequencies of positions within the apparent competition motif varied with degree, trophic level, and their interaction.}
%                 \label{appcomp_DT}
%             \end{figure}

%             \begin{figure}[h!]
%                 \centering
%                 \includegraphics[width=.75\textwidth]{figures/positions_byTL_Directcompetition.eps}
%                 \caption{The frequencies of positions within the direct competition motif varied with degree, trophic level, and their interaction.}
%                 \label{dircomp_DT}
%             \end{figure}

%             \begin{figure}[h!]
%                 \centering
%                 \includegraphics[width=.75\textwidth]{figures/positions_byTL_Three-specieschain.eps}
%                 \caption{The frequencies of positions within the three-species chain motif varied with degree, trophic level, and their interaction.}
%                 \label{chain_DT}
%             \end{figure}

%         \subsection*{Network size and connectance}

%             % Many of these are significant, but the slopes tend to be very shallow. TL and degree are more interesting - this is DEFINITELY SI.
%             The frequency of the top position in the apparent competition motif did not vary strongly with network size or connectance (Table~\ref{SC_tab}).
%             The frequency of the bottom position in the apparent competition motif decreased with increasing connectance but did not vary strongly vary with network size.

%             The frequency of the top position in the direct competition motif increased slightly with the interaction of network size and connectance (i.e., in large and highly-connected networks).
%             The frequency of the bottom position in the direct competition motif was low in all networks.

%             The frequency of all three positions in the omnivory motif showed similar relationships with network size and connectance.
%             Frequencies of these motifs increased with increasing connectance and were not strongly affected by network size.
%             None of the positions in the three-species chain varied strongly with network size or connectance.

%             \begin{table}[h!]
%                 \caption{Coefficients ($\beta$) and $p$-values for linear regressions of the frequency of each motif position against network size, connectance, and their interaction.}
%                 \label{SC_tab}
%                 \footnotesize
%                 \begin{tabular}{c c | c c | cc | cc |}
%                 Motif & Position & \multicolumn{2}{c}{Size} & \multicolumn{2}{c}{Connectance} & \multicolumn{2}{c}{Interaction} \\
%                 & & $\beta$ & $p$ & $\beta$ & $p$ & $\beta$ & $p$ \\
%                 \hline
%                 Chain & bottom & -4.63$\times10^{-5}$ & \textless0.001 & -0.108 & \textless0.001 & -1.94$\times10^{-4}$ & 0.035 \\
%                 Chain & middle & 1.68$\times10^{-4}$ & \textless0.001 & -0.0530 & \textless0.001 & -0.00212 & \textless0.001 \\
%                 Chain & top & 1.00$\times10^{-4}$ & \textless0.001 & 0.0304 & 0.015 & -8.79$\times10^{-4}$ & \textless0.001 \\
%                 \hline
%                 Omnivory & bottom & 2.00$\times10^{-4}$ & \textless0.001 & 0.575 & \textless0.001 & -0.00240 & \textless0.001 \\
%                 Omnivory & middle & 1.32$\times10^{-4}$ & \textless0.001 & 0.551 & \textless0.001 & -0.00215 & \textless0.001 \\
%                 Omnivory & top & 6.42$\times10^{-5}$ & \textless0.001 & 0.448 & \textless0.001 & -0.00115 & \textless0.001 \\
%                 \hline
%                 App. comp. & bottom & -4.74$\times10^{-4}$ & \textless0.001 & -0.663 & \textless0.001 & 0.00224 & \textless0.001 \\
%                 App. comp. & top & 2.12$\times10^{-5}$ & 0.183 & -0.112 & \textless0.001 & -4.47$\times10^{-4}$ & \textless0.001 \\
%                 \hline
%                 Dir. comp. & bottom & 7.94$\times10^{-5}$ & \textless0.001 & -0.0663 & \textless0.001 & 0.00101 & \textless0.001 \\
%                 Dir. comp. & top & -2.45$\times10^{-4}$ & \textless0.001 & -0.602 & \textless0.001 & 0.00610 & \textless0.001 \\
%                 \hline
%                 \end{tabular}
%                 \end{table}

%             \begin{figure}[h!]
%                 \centering
%                 \includegraphics[width=.75\textwidth]{figures/positions_bySC_Omnivory.eps}
%                 \caption{The frequencies of positions within the omnivory motif varied with degree, trophic level, and their interaction.}
%                 \label{omnivory_SC}
%             \end{figure}

%             \begin{figure}[h!]
%                 \centering
%                 \includegraphics[width=.75\textwidth]{figures/positions_bySC_Apparentcompetition.eps}
%                 \caption{The frequencies of positions within the apparent competition motif varied with degree, trophic level, and their interaction.}
%                 \label{appcomp_SC}
%             \end{figure}

%             \begin{figure}[h!]
%                 \centering
%                 \includegraphics[width=.75\textwidth]{figures/positions_bySC_Directcompetition.eps}
%                 \caption{The frequencies of positions within the direct competition motif varied with degree, trophic level, and their interaction.}
%                 \label{dircomp_SC}
%             \end{figure}

%             \begin{figure}[h!]
%                 \centering
%                 \includegraphics[width=.75\textwidth]{figures/positions_bySC_Three-specieschain.eps}
%                 \caption{The frequencies of positions within the three-species chain motif varied with degree, trophic level, and their interaction.}
%                 \label{chain_SC}
%             \end{figure}

%     \clearpage

% \clearpage

% \section*{SJ: Proportion of basal resources in the network}

%     \subsection*{Basal resources and mean persistence}

%         We fit a linear regression relating consumer persistence to network size, connectance, disturbance, the proportion of the network made up of basal resources (pBasal), and all interactions between them. All terms were significant, indicating that the pBasal affects a consumer's probability of persistence but that this effect depends upon other aspects of network structure.
%         As with the effects of network size and connectance, however, the effects of pBasal were small in the range of networks we considered here (Fig.~\ref{fig:lm_BCS}).
%         % Table is from basal_models.Rdata: BCper, if we want it.

%         \begin{figure}[h!]
%             \centering
%             \includegraphics[height=0.75\textheight]{figures/persistence_vs_BSC_lm.eps}
%             \caption{Persistence decreased strongly with increasing probability of disturbance to basal resources, while the proportion of the network made up of basal resources, network size, connectance, and their interactions had smaller effects. We show relationships with persistence for low disturbance ($\pi$=0.1) and high disturbance($\pi$=0.5). For each predictor other than disturbance, we show the relationship between the focal predictor and persistence for combinations of the lowest and highest observed values of the other predictors. For proportion of basal resources, these values were 0.02 and 0.5; for network size, 50 and 100; and for connectance, 0.02 and 0.18.}
%             \label{fig:lm_BCS}
%         \end{figure}

%     \subsection*{Motif profiles and proportion of basal resources}

%         We repeated our analyses testing whether motif profiles were related to network size and connectance, replacing network size with the proportion of the network made up by basal resources (pBasal).
%         A network's overall motif profile was significantly related to pBasal ($F_{1,2996}$=196, $p$\textless0.001, connecance ($F_{1,2996}$=10.0, $p$\textless0.001), and their interaction ($F_{1,2996}$=27.3, $p$\textless0.001).
%         However, this result could be a false positive as the dispersion of motif profiles was non-homogeneous across connectance ($F_{4,2995}$=5.17, $p$\textless0.001), pBasal ($F_{128,2871}$=19.0, $p$\textless0.001), and their interaction ($F_{291,2708}$=13.4, $p$\textless0.001).
%         Distance from group centroids was low for networks with extremely low or extremely high pBasal, perhaps because there were few such networks.
%         The ranges and median distances from group centroids was similar across levels of connectance.

%         Considered separately, the proportions of all four motifs in a network's motif profile were related to the proportion of basal resources in a network (Fig.~\ref{fig:Blms}).
%         The proportion of the three-species chain decreased in networks with more basal resources ($\beta_{pBasal}$=-0.080, $p$\textless0.001) and networks with higher connectance ($\beta_{C}$=0.498, $p$\textless0.001).
%         The interaction between pBasal and connectance was not significant ($\beta_{pBasal:C}$=0.0554, $p$=0.766).

%         The proportion of the omnivory motif decreased in networks with more basal resources ($\beta_{pBasal}$=-0.198, $p$\textless0.001), increased in networks with higher connectance ($\beta_{C}$=0.895, $p$\textless0.001), and increased most strongly in networks with high connectance and many basal resources ($\beta_{pBasal:C}$=2.53, $p$\textless0.001).

%         The proportion of the apparent competition motif increased with in networks with more basal resources ($\beta_{pBasal}$=0.202, $p$\textless0.001) and decreased in more connected networks ($\beta_C$=-0.596, $p$\textless0.001). 
%         The interaction between the proportion of basal resources and connectance was not significant ($\beta_{pBasal:C}$=0.258, $p$\textless0.001).

%         The proportion of the direct competition motif increased in networks with more basal resources ($\beta_{pBasal}$=0.0759, $p$\textless0.001) and more-connected networks ($\beta_C$=0.199, $p$\textless0.001), but decreased with the interaction between them ($\beta_{pBasal:C}$=-2.85, $p$\textless0.001).

%         \begin{figure}
%             \centering
%             \includegraphics[width=.75\textwidth]{figures/motif_proportion_basal_lms.eps}
%             \caption{Mean proportion of each motif in a network's motif profile varied with connectance and the proportion of basal resources in the network. For direct competition and omnivory, the frequency of the motif also varied with the interaction between connectance and basal resources.}
%             \label{fig:Blms}
%         \end{figure}

%     \subsection*{Motif participation and proportion of basal resources}

%         We fit regressions relating the proportion of each motif in a consumer's motif participation vector to the proportion of basal resources in the network (pBasal), connectance, and the interaction between them.
%         Participation in all four motifs was significantly related to pBasal, connectance, and the interaction between them (Fig.~\ref{fig:basal_participation}).
%         These interactions were strongest for apparent and direct competition.

%         \begin{figure}
%             \centering
%             \includegraphics[height=.65\textheight]{figures/participation_vs_BC.eps}
%             \caption{The proportion of each motif in a consumer's participation vector varied with the proportion of the network made up by basal resources, connectance, and their interaction.}
%             \label{fig:basal_participation}
%         \end{figure}

%         Participation in the three-species chain increased in networks with more basal resources ($\beta_{pBasal}$=0.0447, $p$\textless0.001), decreased in more-connected networks ($\beta_C$=-0.296, $p$\textless0.001), and decreased with the interaction between them ($\beta_{pBasal:C}$=-0.525, $p$\textless0.001).
%         Participation in the omnivory motif decreased in networks with more basal resources ($\beta_{pBasal}$=-0.198, $p$\textless0.001), increased in more-connected networks ($\beta_C$=0.701, $p$\textless0.001), and increased strongly with the interaction between them ($\beta_{pBasal:C}$=2.81, $p$\textless0.001).
%         Participation in the apparent competition motif decreased in networks with more basal resources and more-connected networks ($\beta_{pBasal}$=-0.0825, $p$\textless0.001 and $\beta_C$=-0.846, $p$\textless0.001) but increased strongly with the interaction between them ($\beta_{pBasal:C}$=1.66, $p$\textless0.001).
%         Participation in the direct competition motif increased in networks with more basal resources and more-connected networks ($\beta_{pBasal}$=0.235, $p$\textless0.001 and $\beta_C$=0.442, $p$\textless0.002), but decreased strongly with the interaction between them ($\beta_{pBasal:C}$=-3.94, $p$\textless0.001).

%     \subsection*{Frequencies of positions and proportion of basal resources}

%         As with the frequencies of each motif in a species' motif participation vector, the frequencies of positions in a species' motif role varied with the proportion of basal resources in the network, connectance, and their interaction (Table~\ref{BC_tab}).

>>>>

\documentclass[12pt]{article} 
\usepackage{amsmath} 
\usepackage[dvips]{graphicx}
\usepackage{multirow} 
\usepackage{geometry} 
\usepackage{pdflscape}
\usepackage[labelfont=bf]{caption} 
\usepackage{setspace}
\usepackage[running]{lineno} 
% \usepackage[numbers,sort]{natbib}
\usepackage[numbers]{natbib} 
\usepackage{array}
\usepackage[table]{xcolor}
\usepackage{xr}

@@

\title{Species motif participation provides unique information about species risk of extinction}

\author{Alyssa R. Cirtwill$^{1*\dagger}$, Anna \r{A}kesson$^{2*}$, Kate L. Wootton$^{3}$, Anna Ekl\"{o}f$^{2}$} 
\date{
\small$^1$Spatial Foodweb Ecology Group\\
Research Centre for Ecological Change\\
Organismal and Evolutionary Biology Research Programm\\
Faculty of Biological and Environmental Sciences\\
University of Helsinki, Helsinki, Finland\\
\medskip
\small$^2$Department of Theoretical Biology, Chemistry, and Physics\\ 
Link\"{o}ping University, Link\"{o}ping, Sweden\\
\medskip
\small$^3$ BioFrontiers Institute\\
University of Colorado, Boulder, Boulder, U.S.A.\\
\medskip
* Joint first authors\\
\medskip
$^\dagger$ Corresponding author\\
}

@@

\section*{Table of contents}
    {\footnotesize
    \begin{spacing}{1.0}

@@

    \end{enumerate}    
\end{spacing}
}
\clearpage

@@

        We use Bayesian network modelling to calculate the persistence of each species in our simulated networks. 
        A Bayesian network represents a graphical model for probabilistic relationships among a set of variables. 
        In our ecological setting, the nodes of the network represent species and links represent feeding interactions between species.
        In network-theory terms, the species are Bernoulli random variables and the links represent conditional dependencies among these variables. 
        Each node has a set of local probability dependencies that are conditional on the state of the parent (resource) nodes. 
        Since the nodes are species, their state can be either extant or extinct.
        The extinction probability of each species is therefore a function of the state of its resources and also depends on a baseline probability of extinction ($\pi$) representing risk of extinction due to factors unrelated to network structure.
        In our baseline scenario, basal resources had a baseline extinction probability of $\pi = 0.1$. 
        We also simulated a set of scenarios where basal resources had a higher baseline probability of extinction ($\pi$). 
        These scenarios reflect threats to basal resources such as increased temperatures resulting in droughts, making some species more vulnerable to extinction.
        $\pi$ ranged between $0.1-0.5$, in steps of $0.08$. 
        The highest disturbance level, $\pi = 0.5$, corresponds to basal species having a 50\% risk of going extinct. 
        Consumer species retained $\pi=0.1$ in all cases.

@@

        As consumer species depend on their resources, we start by determining the status of each primary producer in the network.
        Systematically, we use Equation \ref{betafunc} to calculate each species' probability of extinction $P(\lnot i|f)$, and then perform Bernoulli trials to determine whether the species goes extinct or not. 
        Imagine a basal species \textit{A} with extinction probability $P(\lnot A) = \pi$ where $\pi = 0.1$. 
        We draw a random number $r$ from a uniform distribution between $0-1$.
        The basal species would be considered extinct if $r < 0.1$ and extant otherwise.

@@

        We simulated a suite of food webs based on the probabilistic niche model, which assigns predator-prey links based on the body-mass ratios between individuals of different species~\citep{Williams2000,Delmas2017}. 
        The meso-scale structure of niche-model networks closely mimics that of empirical food webs~\citep{Stouffer2007}. 
        To ensure that we captured a variety of realistic community sizes and structures, we generated networks ranging between 50 and 100 species (in steps of 10) with connectance values between 0.02 and 0.2 (in steps of 0.02). 
        The range of network sizes was chosen to reflect moderately well-sampled empirical webs while working within our computational limits, while the range of connectance values was chosen to cover that observed in most empirical food webs~\citep{Dunne2002e}.
        We generated a total of 100 networks with each combination of parameters, for a total of 6000 networks. 
        All networks were generated using the function ``nichemodel'' within the Julia language package \emph{BioEnergeticFoodWebs}~\citep{bioenergeticfw,Delmas2017}. 
        If a simulated network contained any disconnected species (species without predators or prey) or disconnected components (a group of species connected among themselves but not to the rest of the network), the network was rejected and a new network simulated.
        Finally, networks where the path lengths between each species and a basal resource could not be resolved (i.e., trophic levels were undefined) were rejected and new networks simulated.

        As a further check that our simulated networks are realistic (e.g., could plausibly retain all species), we simulated community dynamics for 500 timesteps and discarded any networks which did not retain all species. 
        We simulated dynamics using the function ``simulate'' from the Julia language package \emph{BioEnergeticFoodWebs}~\citep{bioenergeticfw,Delmas2017}.
        This function uses Lotka-Volterra predator-prey models including density dependence and type 2 functional responses for all species (please see~\citet{Delmas2017} for full details).
        All non-basal species were designated as vertebrates to ensure a good match between metabolic and predator-prey body-mass ratio values. Metabolic rates in the Lotka-Volterra model are based on each species' body mass (i.e. mass of a single individual). 
        We assigned relative body masses based on each species' trophic level, which was, in turn, calculated based on the food-web structure provided by the niche model. 
        After basal species were assigned a body mass of 1, we used a predator-prey body-mass ratio of 3.065 to calculate the relative body masses of higher trophic levels. 
        We selected this ratio based on the estimate for vertebrates (averaged across ecosystem and metabolic types) in~\citet{Brose2006}. 
        We excluded reported body-mass ratios for invertebrates as these could include parasites and parasitoids, which are generally smaller than their prey, and because interactions among vertebrates are better represented in the food-web literature than interactions involving invertebrates.

        The persistence of each species in our simulated networks also depends on its population biomass. 
        We randomly assigned initial population biomasses (i.e. cumulative biomass across all individuals of a species) for each species from a uniform distribution [0,1]. 
        Note that population biomasses and individual body masses are not calculated on the same scale. 
        We then simulated community dynamics for 1000 time steps to obtain an `equilibrium' community. 
        To ensure that species did not `recover' from unrealistically low biomasses during the simulation, we considered a species extinct if it dropped below an arbitrary threshold biomass of 1$\times10^{-5}$. 
        We rejected any network where one or more species dropped below this biomass threshold. 
        Consumers were assumed to have no preferences such that the consumption rate $w_{ij}$ of predator $i$ eating prey $j$ is equal to $\frac{1}{n}$, where $n$ is the number of prey for predator $i$. 
        If the network did not retain all species for 1000 time steps, a new set of initial population biomasses was applied and the simulation repeated.
        If a set of biomasses where all species persist still had not been reached after 100 sets of randomly-assigned initial biomasses, we discarded the network and simulated another to replace it.

@@

        A Bayesian food web depicts probabilistic relationships among a set of species where each species' probability of persistence depends upon the probabilities of its resource species persisting~\citep{Jensen_Nielsen,Eklof2013}. 
        All persistence calculations therefore begin by determining the status of the basal resources (who do not depend on other species).
        Calculations continue in a strictly bottom-up manner with primary consumers (who depend only on basal resources), then herbivores, and so on up the network.

@@

        \begin{figure}
            \centering
            \includegraphics[width=\textwidth]{figures/motif_profiles_participation_vs_empirical.eps}
            \caption{\textbf{A)} Proportions of acyclic motifs in the motif profiles of empirical and simulated networks. Mean proportions ($\pm$SD) in the simulated networks are indicated in black, with the range of proportions in the simulated networks indicated by black boxes. \textbf{B)} Proportions of acyclic motifs in the motif participation vectors of consumers in simulated and empirical networks.}
            \label{empirical_compare}
        \end{figure}

@@

    \begin{figure}[hb!]
        \centering
      \includegraphics[width=\textwidth]{figures/heatmap_CS_BPcompare.eps}
       \caption{Heat maps of average consumer persistence in networks with varying connectance (y-axis) and network size (x-axis) at two levels of disturbance to basal resources. Persistence decreases with darker color in the heat map. When basal species are not disturbed, all species experience a baseline extinction risk of $\pi = 0.1$. When basal species are highly disturbed, their baseline extinction risk rises to $\pi = 0.5$ while the baseline risk of consumers is not changed.}
       \label{fig:heatmap_CS}
    \end{figure}

@@

\bibliographystyle{vancouver} 
\bibliography{anna_bib_new} % Abbreviate journal titles.
